---
format: 
  html:
    page-layout: full
  pdf: default
---

# Literature Review Data

This table provides a tabular overview of the studies surveyed as part of this study's literature review on textbook research examining describing and/or evaluating the language of English textbooks designed for English L2 learners in various instructional settings.

It presents the results of this non-exhaustive survey of Textbook English studies published over the past four decades, summarising some of the key information on each study, including its main language focus, methodological approach, information on the textbooks investigated, and, if applicable, on any reference corpora used. Empty cells represent fields for which no information was published. The table is fully searchable and filterable. You can adjust the widths of the individual columns to best fit your screen size.

This list is intended to be a dynamic resource that will grow over time. If you would like to contribute any studies to the table, please either fork [the corresponding CSV file in the repository](https://github.com/elenlefoll/TextbookEnglish/blob/main/LitReviewTable.csv) or send me an [e-mail](mailto:elefoll@uni-koeln.de?subject=%5BLitReviewTable%5D%20Suggestions) with the corresponding details of the studies that you would like to add.

```{r date, include=FALSE}

file <- here("data", "LitReviewTable.csv")

```

**This data displayed on this page was last updated on `r format(file.info(file)$ctime, '%d %B %Y')`.**

```{r setup, include=FALSE}
library(dplyr)
library(here)
library(reactable) 
options(scipen=999)
```

```{r import, include=FALSE}

table <- read.csv(here("data", "LitReviewTable.csv"))

```

```{r add url, include = FALSE}

# Adding working links
html.table <- table %>% 
  mutate(Reference = stringr::str_replace(Full.bibiographic.reference, "doi:|DOI:", "https://doi.org/")) %>% 
  mutate(Reference = stringr::str_replace_all(
    string = Reference, 
    pattern = "(((f|ht)tp(s)?://)[-a-zA-Z?-??-?()0-9@:%_+.~#?&;//=]+)",
    replacement = "<a href='\\1' target=\"_blank\" rel=\"noopener noreferrer\" > \\1</a>")) %>% 
  select(-Full.bibiographic.reference) %>% 
  rename(Method = Methodology) %>% 
  rename(Ref.corpus = Reference.corpus)
  
#View(html.table)

```

```{r display, echo=FALSE}

library(webshot)

reactable(html.table, 
          filterable = TRUE, searchable = TRUE, bordered = FALSE, 
          striped = TRUE, highlight = TRUE, resizable = TRUE,
          defaultPageSize = 10, showPageSizeOptions = TRUE, 
          showSortable = TRUE, outlined = TRUE,
          pageSizeOptions = c(10, 25, 50, 100), 
          defaultSortOrder = "desc",
            columns = list(
            Year = colDef(defaultSortOrder = "asc", minWidth = 70),
            Title = colDef(minWidth = 200),
            Setting = colDef(minWidth = 50),
            TxB.level = colDef(minWidth = 100),
            Nb.TxBs = colDef(minWidth = 60),
            TxB.pub.date = colDef(minWidth = 50), 
            Method = colDef(minWidth = 150),
            Ref.corpus = colDef(minWidth = 150),
            Reference = colDef(html = TRUE, minWidth = 400)
            )
) 

```

The raw data can be downloaded as a comma-separated file from [the project GitHub repository](https://github.com/elenlefoll/TextbookMDA).
